What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-07 android-stacks.tex
Sections 1 and 2

To answer this question, we mined the following information of every extracted
stack trace: the top level exception being signaled, its signaler (the first
method signature frame after the exception), the root exception (the exception
which initiated the stack trace, defined on the bottom of the stack), and the
root signaler. The intermediate elements that compose a stack trace as shown in
Figure 1 (i.e., the intermediate causes and its execution stacks) were not be
considered in the analysis.
%~\todo{Move this to the previous section? IMO, It is better to keep all facts
%about how we did the research in one place}

Based on the assumption that, regardless the issue type every exception stack
trace contains relevant information concerting the exception structure of
projects analyzed we opted for not restricting the analysis only on defect
issues~\footnote{We conducted the same analysis on the defect issues and the top
exceptions are similar to the ones mentioned on defect issues. Due to space
limitation we limit to present the general analysis here. More detailed analysis
on the defect issues can be found here:
\url{www.dimap.ufrn.br/~roberta/icsm2014}}.

@@

Table~\ref{tab:toptenjava} illustrates the top 10 most common root causes which
represents more than 50\% of the analyzed set. The table illustrates  (i) the
number of times each exception appeared as the root cause of a stack trace
(\#occurrences), (ii) the number of distinct projects reporting such causes on
stack traces, and (iii) the exception popularity. The exception popularity
metric was defined in this work as follows: \emph{popularity = number of
projects mentioning the exception in at least one stack/all projects that
contain at least one stack}.

\begin{table}
\begin{tabular}{lrrrr}
    \hline
    \bfseries{Exception} &  \multicolumn{2}{c}{\bfseries{Occurrences}} &  \multicolumn{2}{c}{\bfseries{Projects}}\\
    & \bfseries{\#} &  \bfseries{\%} & \bfseries{\# } & \bfseries{\% } \\
    \hline
java.lang.NullPointerException             & 7578 & 26,31\% & 1671 & 44,47 \\
java.lang.IllegalArgumentException         & 1573 & 5,46\%  & 670  & 17,83\% \\
java.lang.ClassNotFoundException           & 1203 & 4,18\%  & 514  & 13,68\% \\
java.lang.RuntimeException                 & 1093 & 3,80\%  & 411  & 10,94\% \\
java.lang.IllegalStateException            & 964  & 3,35\%  & 431  & 11,47\% \\
java.lang.ClassCastException               & 886  & 3,08\%  & 408  & 10,86\% \\
java.lang.ArrayIndexOutOfBoundsException   & 834  & 2,90\%  & 345  & 9,18\% \\
java.lang.NoSuchMethodError                & 822  & 2,85\%  & 369  & 9,82\% \\
java.io.IOException                        & 656  & 2,28\%  & 337  & 8,97\% \\
java.lang.OutOfMemoryError                 & 601  & 2,09\%  & 235  & 6,25\% \\
%java.lang.NoClassDefFoundError            & 554  & 1,92\%  & 299  & 7,96\% \\
%java.util.ConcurrentModificationException & 465  & 1,61\%  & 180  & 4,79\% \\
%java.lang.IndexOutOfBoundsException       & 445  & 1,55\%  & 205  & 5,46\% \\
\hline
All	 &	23956 &	100\%	& 3758 & 100\%	 \\
\hline
  \end{tabular}
\caption{Top 10 most popular exceptions on issues}
\label{tab:toptenjava}
\end{table}

We can observe that the NullPointerException was the one with the higher number
of occurrences (22,5\% of all stacks were caused by NullPointerExceptions) and
the one which had the highest popularity across projects. Actually, most of the
exceptions in the TOP 10 list (except java.lang.IllegalArgumentException,
java.io.IOException and java.lang.RuntimeException) are exceptions implicitly
signaled by Java Virtual Machine, when an expression violates the normal
semantics of the Java programming language (e.g. ArrayIndexOutOfBounds) or due
to an internal error or resource limitation (OutOfMemoryError).

The high prevalence of NullPointerExceptions, and the other implicitly-thrown
exceptions  is aligned with the findings other works [Sunghun et al.][Gordon
fraser and][Jcrasher].[Android-predict]~\todo{refs}. For instance, Sunghun et
al. showed that in Eclipse bug report system 38\% the bugs related to exception
handling were caused by NullPointerException; other works on robustness testing
[JCrasher, Evol;suite, Amiya]~\todo{refs} showed that most of the automatically
detected bugs were due to NullPointerExceptions and implicitly-signaled of Java
environment (as the ones found in this study).

@@

As detailed in Section III, 482 Android projects were carefully selected to
enable a deaper investigation of stack trace information, by leveraging stack
trace information with the source code and bytecode analysis. For this subset of
Android projects we could extract every package related to each exception
signaler, and based on this information we could classify every stack trace
based on its root signaler as presented in Table~\ref{signalers}.

@@

For each exception we could discover: the number of times it occurred on stacks,
the number of distinct projects such stacks were defined (and consequently its
popularity) and the of times it was signaled by different signaler types. We
could calculate the popularity of each root cause, considering the number of
distinct projects they were found and the whole set of projects analyzed with at
least one stack trace (482). Table~\ref{toptenandroid} and
Figuere~\ref{androidsignaler} presents the mined data.

@@

%\textbf{
Coding errors are the most common root causes for most types of signalers. We
could observe that the NullPointerException is still the exception with higher
number of occurrences. The NullPoiterExceptions are mainly signaled inside
Android platform and inside application code. Although we also find
NullPointerExceptions being signaled from third-party libraries. Regarding
reusable code, there is no consensus if it s a good or bad practice to re-throw
NullPointerException. Some prefer to encapsulate such exception on
IllegalArgumentException. But others~\cite{joshuabloch} argue that the
NullPointerException  explicits the cause of the problem and hence should not be
wrapped.

%commented only for submission:
\textbf{We could observe in this study that the NullPointerException and other
other implicitly signaled exceptions represented most of the exceptions reported
on issues in both Java and Android subset. For such exceptions - which represent
programming bugs or resource limitations - there is usually no proper handling
besides presenting an error message to the user and restarting the application -
only high fault tolerant systems need to provide solutions to handle them [REF:
Java Specification].}

>>>>

To answer this question, we mined from every extracted stack trace the top level
exception being signaled, its signaler (the first method signature frame after
the exception), the root exception (the exception which initiated the stack
trace, defined on the bottom of the stack), and the root signaler. The
intermediate elements that compose a stack trace as shown in
Figure~\ref{fig:wrapping} (i.e., the intermediate causes and its execution
stacks) were not be considered in the analysis.
%~\todo{Move this to the previous section? IMO, It is better to keep all facts
%about how we did the research in one place}

Based on the assumption that regardless the issue type, every exception stack
trace contains relevant information concerning the exception structure of
projects analyzed we opted for not restricting the analysis only on defect
issues~\footnote{We conducted the same analysis on the defect issues and the top
exceptions are similar to the ones mentioned on defect issues. Due to space
limitation we limit to present the general analysis here. More detailed analysis
on the defect issues can be found here:
\url{www.dimap.ufrn.br/~roberta/icsm2014}}.

@@

Table~\ref{tab:toptenjava} illustrates the top 10 most common root causes which
represents more than 50\% of the analyzed set. The table illustrates  i) the
number of times each exception appeared as the root cause of a stack trace
(\#occurrences), ii) the number of distinct projects reporting such causes on
stack traces, and iii) the exception popularity among projects.

\begin{table}
\begin{tabular}{lrrrr}
    \hline
    \bfseries{Exception} &  \multicolumn{2}{c}{\bfseries{Occurrences}} &  \multicolumn{2}{c}{\bfseries{Projects}}\\
    & \bfseries{\#} &  \bfseries{\%} & \bfseries{\# } & \bfseries{\% } \\
    \hline
java.lang.NullPointerException             & 7578 & 26,31\% & 1671 & 44,47 \\
java.lang.IllegalArgumentException         & 1573 & 5,46\%  & 670  & 17,83\% \\
java.lang.ClassNotFoundException           & 1203 & 4,18\%  & 514  & 13,68\% \\
java.lang.RuntimeException                 & 1093 & 3,80\%  & 411  & 10,94\% \\
java.lang.IllegalStateException            & 964  & 3,35\%  & 431  & 11,47\% \\
java.lang.ClassCastException               & 886  & 3,08\%  & 408  & 10,86\% \\
java.lang.ArrayIndexOutOfBoundsException   & 834  & 2,90\%  & 345  & 9,18\% \\
java.lang.NoSuchMethodError                & 822  & 2,85\%  & 369  & 9,82\% \\
java.io.IOException                        & 656  & 2,28\%  & 337  & 8,97\% \\
java.lang.OutOfMemoryError                 & 601  & 2,09\%  & 235  & 6,25\% \\
%java.lang.NoClassDefFoundError            & 554  & 1,92\%  & 299  & 7,96\% \\
%java.util.ConcurrentModificationException & 465  & 1,61\%  & 180  & 4,79\% \\
%java.lang.IndexOutOfBoundsException       & 445  & 1,55\%  & 205  & 5,46\% \\
\hline
All	 &	23956 &	100\%	& 3758 & 100\%	 \\
\hline
  \end{tabular}
\caption{Top 10 most popular exceptions on issues in Java projects}
\label{tab:toptenjava}
\end{table}

We can observe that the NullPointerException was the one with the higher number
of occurrences (26,3\% of all stacks were caused by NullPointerExceptions) and
the one which had the highest popularity across projects. Actually, most of the
exceptions in the top 10 list (except java.lang.IllegalArgumentException,
java.io.IOException and java.lang.RuntimeException) are exceptions implicitly
thrown by Java Virtual Machine, when an expression violates the normal
semantics of the Java programming language (e.g. ArrayIndexOutOfBounds) or due
to an internal error or resource limitation (OutOfMemoryError).

The high prevalence of NullPointerExceptions, and the other implicitly-thrown
exceptions  is aligned with the findings other works~\cite{kim2013predicting,fraser20131600,csallner2004jcrasher}. For instance, Sunghun et
al.~\cite{kim2013predicting} showed that in Eclipse bug report system 38\% the bugs related to exception handling were caused by NullPointerException; other works on robustness testing~\cite{maji2012empirical,csallner2004jcrasher}
showed that most of the automatically
detected bugs were due to NullPointerExceptions and implicitly-signaled of Java
environment (as the ones found in this study).

@@

As detailed in Section~\ref{sec:android}, 482 Android projects were carefully
selected to enable a deeper investigation of stack trace information, by
leveraging stack trace information with the source code and bytecode analysis.
For this subset of Android project, we could extract every package related to
each exception signaler, and based on this information we could classify every
stack trace based on its root signaler as presented in Table~\ref{tab:signalers}.

@@

For each exception we could discover the number of times it occurred on stacks,
the number of distinct projects such stacks were defined (and consequently its
popularity) and the of times it was signaled by different signaler types. We
could calculate the popularity of each root cause, considering the number of
distinct projects they were found and the whole set of projects analyzed with at
least one stack trace (482). Table~\ref{tab:toptenandroid} and
Figure~\ref{fig:androidsignaler} presents the mined data.

@@

Coding errors are the most common root causes for most types of signalers. We
could observe that the NullPointerException is still the exception with higher
number of occurrences. The NullPoiterExceptions are mainly signaled inside
Android platform and inside application code, although we also find
NullPointerExceptions being signaled from third-party libraries. Regarding
reusable code, there is no consensus if it is a good or bad practice to re-throw
NullPointerException. Some prefer to encapsulate such an exception on
IllegalArgumentException, while others~\cite{bloch2008effective} argue that the
NullPointerException makes the cause of the problem explicit and hence 
should not be wrapped.

%commented only for submission:
We could observe in this study that the NullPointerException and other
other implicitly signaled exceptions represented most of the exceptions reported
on issues in both Java and Android subset. For such exceptions, which represent
programming bugs or resource limitations, there is usually no proper handling
besides presenting an error message to the user and restarting the application -
only high fault tolerant systems need to provide solutions to handle them.

\noindent \fbox{
\begin{minipage}{0.96\columnwidth} 
\emph{RQ2: Most exceptions are due to programming mistakes
or generally irrecoverable errors.\todo{Revise this}}
\end{minipage}}
